---
title: ANOVA in R
subtitle: Cheatsheet
date: today
lightbox: true
---

```{r setup}
#| include: false

if (!require("Require")) {
    install.packages("Require")
    require("Require")
}
Require(c(
    car, emmeans, gt, gtsummary, rstatix, tidyverse, palmerpenguins
))
possums <- readxl::read_excel("possum_bw.xlsx")
```


::: {.grid}
::: {.g-col-6}
{{< include _license.qmd >}}

::: {.callout-note collapse="false"}
## Assumed knowledge
- You know how to install and load packages in R.
- You know how to import data into R.
- You recognise data frames and vectors.
- **You know what an ANOVA test is and when to use it. **
:::
:::
::: {.g-col-6}
{{< include _tidy.qmd >}}

::: {.callout-warning}
## Data
For this cheatsheet we will use data from the penguins dataset from the `palmerpenguins` package. You may need to install this package:

```{r}
#| eval: false
install.packages("palmerpenguins")
```

:::
:::
:::


## About
ANOVA summaries are often used when the model's predictors/explanatory variables are **categorical**. This cheatsheet shows you how to perform ANOVA techniqus in R.

ANOVA answers questions such as:

- Are there differences in plant height between these two locations? (Yes, similar to a two-sample *t*-test).
- Is location a significant predictor of species diversity?
- Is the mean height of kangaroos different between different species?

[**Interpretation of the results is not covered here.**]{style="color:red"}

### R packages used
`tidyverse`, `car`, `emmeans`, `gt`, `gtsummary`, `rstatix`, `palmerpenguins`



## Similarity between ANOVA and linear regression
All of the below are equivalent ways to perform an ANOVA test in R. The first method is the most common but method 4 is most flexible. Note that [native R pipes (`|>`)](https://posit.co/wp-content/themes/Posit/public/markdown-blogs/rstudio-v1-4-update-whats-new/index.html#new-native-pipe-operator) are used in the examples below.

```{r}
#| eval: false

## ANOVA
aov(body_mass_g ~ species, data = penguins) |> summary() # method 1

## Using GLM (multiple ways)
lm(body_mass_g ~ species, data = penguins) # method 2; same p-values
lm(body_mass_g ~ species, data = penguins) |> anova() # method 3
lm(body_mass_g ~ species, data = penguins) |> car::Anova() # method 4
```

## ANOVA types
All ANOVA tables are created using the `car::Anova()` function, which are more flexible and can create type II and type III ANOVA tables rapidly (useful for unbalanced designs and/or with interactions).

::: panel-tabset
## One-way ANOVA
```{r}
#| eval: true
fit01 <- lm(body_mass_g ~ species, data = penguins)
car::Anova(fit01) # prints the ANOVA table
 
```

## Two-way ANOVA
```{r}
#| eval: true
fit02 <- lm(body_mass_g ~ species + sex, data = penguins)
car::Anova(fit02) # prints the ANOVA table
```

## Two-way ANOVA with interaction
```{r}
#| eval: true
fit03 <- lm(body_mass_g ~ species * sex, data = penguins)
car::Anova(fit03) # prints the ANOVA table
```
:::


## Assumptions

```{r}
#| fig-height: 6
par(mfrow = c(2, 2)) # creates a 2x2 grid for plots
plot(fit01)
```


## Interpreting effects using estimated marginal means


::: panel-tabset
## One-way ANOVA
The simplest way to report pairwise comparisons is to plot the estimated marginal means (EMMs) and their confidence intervals.
```{r}
#| fig-height: 1.5
emm1 <- emmeans(fit01, ~ species)
plot(emm1, comparisons = TRUE)
```

Reporting of the plots can be supplemented with a table of the estimated marginal means (perhaps in the appendix or as supplementary material).

```{r}
emm1
```

## Two-way ANOVA
As there are two main effects, we need to plot the estimated marginal means for each main effect separately.
```{r}
#| fig-height: 1.5
emm2 <- emmeans(fit02, ~ species)
plot(emm2, comparisons = TRUE)
emm3 <- emmeans(fit02, ~ sex)
plot(emm3, comparisons = TRUE)
```


## Significant interactions
If there is a significant interaction, the main effects should not be interpreted without considering the interaction. Two plots are needed to interpret a two-way interaction.

```{r}
#| fig-height: 1.6
emmip(fit03, sex ~ species)
emmip(fit03, species ~ sex)
```
:::

## Other resources

- It might be worthwhile to use the [`performance`](https://easystats.github.io/performance/) package to assess model fit (including assumptions using `check_model()`).
- I use this a lot: the [`interactions`](https://interactions.jacob-long.com/) package for visualising interactions in GLM models. However it is very technical and not for beginners -- use if you are comfortable with R.
- The `rstatix` package has a useful function called [`anova_summary()`](https://rpkgs.datanovia.com/rstatix/reference/anova_summary.html) to produce *almost publishable* ANOVA tables. You will still need to manipulate it to make it look nice.

